---
title: |-
  How to make an R package
   \vspace{0.5in}
author:
- name: Christoffer Trier Maansson
  affiliation: 
    Department of Clinical Biochemistry, Aarhus University Hospital, Denmark
  email: ctm@clin.au.dk
date: "`r format(Sys.time(), '%d %b %Y')`"
output:
  BiocStyle::html_document:
    toc: yes
    toc_depth: 3
    number_sections: yes
    highlight: haddock
abstract: "This document will demonstrate how to make an
R package. First it will highlight the contents of an R
Package and then move on to explain how a package is 
created, edited, updated and eventually submitted to
for example Bioconductor. The contents of an R package
is located at a GitHub repository and therefore a short
introduction to GitHub is also included. I have created an R package called
**DNAfusion** which can be located at my GitHub repository
https://github.com/CTrierMaansson/DNAfusion. This can serve as an example of
an R package. " 
vignette: |
  %\VignetteIndexEntry{How to make an R package}
  %\VignetteEncoding{UTF-8} 
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    tidy.opts = list(width.cutoff=100),
    tidy = FALSE,
    message = FALSE,
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction to **GitHub**
An R package is first and foremost a collection of R functions created with a 
specific purpose, however the package also needs documentation, tests,
descriptions, and more. All of the files for the R package is kept in a folder,
which is linked to a GitHub repository. This means that all of the code in the 
package is freely available to the public. 

## What is a **GitHub repository**
A **GitHub repository** is an online folder which is linked to
a folder on your local computer. When you make changes to that folder they are 
registered and can be `pushed` to the online repository, updating the content
of the online folder. In order to update the repository the changes first need
to be `commited`, followed by defining a short name for the update and then 
`pushed` to the repository. I strongly recommend using the freely available 
tool GitHub Desktop, which can be used to perform these tasks instead of 
writing the commands in the prompt. 

## **GitHub branches**
Because **GitHub repositories** are accessible to everyone it is frequently 
used in collaborations. This gives the opportunity for people to work on a
project within the same folder without the fear of issues with synchronization.
This is achieved through `branches`. GitHub have a memory of previous versions
of the repository and each time a change is pushed to the repository it is 
pushed to a specific `branch` of the repository. As default, the first 
branch is called `main` (previously called `master`, however
because of black lives matter, this was changed to `main`... True story)
and can be viewed as the stem of a tree. Lets say an individual wants to 
investigate some changes to the repository however they don't want to change
the main repository.Then they simply create a new branch 
(let's call it `alternative`) and push the changes to that branch. Now the 
`main` and `alternative` branches can exist in parallel and be modified 
separately. If `alternative` eventually prove superior `alternative` can be
merged into `main` and now only the `main` branch exist. 

It is important to remember that if more than one person is working on a
project you need to `fetch` the most recent version of the repository before
any changes are made locally and then pushed to the repository. 
This habit of fetching, commiting and pushing quickly because standard.

**For more information regarding GitHub i recommend this tutorial:**

https://www.youtube.com/watch?v=8Dd7KRpKeaE&ab_channel=CoderCoder


# **Contents** of an R Package
The repository must contain 4 folders (man, R, tests, and vignettes) and 3 
documents (DESCRIPTION, NAMESPACE, and README). In addition it can also
contain the folder inst, as well as LICENSE and NEWS documents. 
**All folders and files must be given these exact names!**

## The **R** folder
This folder contains the functions of the package. The functions can either be
in the same R document (.R) or in separate documents.
All functions in R must have help documents which can be read in R using 
either of the two commands:

```{r helper, message=FALSE, results = 'hide', eval = FALSE}
?my_function()

#or

help(my_function)

```

`my_function()` is the name of the function to be included in the package.

`roxygen2` is used to create this documentation and includes so called tags
which are defined using `@`. `roxygen2` will not be explained in this tutorial
however more information can be found at https://r-pkgs.org/man.html. 

## The **man** folder
The man folder contains the documentation of each function of the package in 
separate .Rd files. **Importantly, these files are not made manually**. They 
are created using either:

```{r document, message=FALSE, results = 'hide', eval = FALSE}
if (!require(devtools)) install.packages("devtools")
library(devtools)
devtools::document()

#or
if (!require(roxygen2)) install.packages("roxygen2")
library(roxygen2)
roxygen2::roxygenize()

```

As demonstrated above, the package is created by running other functions. I
**STRONGLY** recommend creating an .R document 
(for example creating_MyPackage.R) **NOT** located in the package folder where 
you can run code which will build, check, install the package among 
other things. But more on that later. 

The .Rd files are automatically put into the man folder. 

## The **tests** folder
This folder is only mandatory if the package needs to be submitted to CRAN or
Bioconductor. The folder includes an .R document called testhat.R which just 
contains 4 lines of code:

```{r testthat, message=FALSE, results = 'hide', eval = FALSE}
if (!require(testthat)) install.packages("testthat")
library(MyPackage)
library(testthat)
test_check("MyPackage")

```

Additionally, tests, contains the folder, testthat, which includes a single .R 
document called test_functions.R

This document uses functions from the `testthat` package to test the code in
the R folder. More specifically, it runs the code and test if the output are
of the expected class. An R package needs a test document which covers the 
code in the R file, meaning most of the code is run during the test. 
The coverage of the tests are investigated using. 

```{r coverage, message=FALSE, results = 'hide', eval = FALSE}
if (!require(covr)) install.packages("covr")
library(covr)
covr::report()

```

It is expected that the coverage should be above 90%. 

For more information regarding tests see:

https://www.r-bloggers.com/2019/11/automated-testing-with-testthat-in-practice/


## The **vignettes** folder
This folder contains a single R markdown document (.rmd) which works as a 
guide for how to use the package. A vignette is typically a .html or .pdf file 
which will walk you through all functions with examples and explains how to 
implement each function. Vignettes can vary a lot in detail however a good 
vignette makes it easier for outsiders to understand the package
and is therefore always appreciated. CRAN and bioconductor have specific
requirements to the contents of the vignette. 

How to create the vignette will not be discussed here, however I recommend 
referring to 

https://github.com/CTrierMaansson/DNAfusion/tree/main/vignettes
to see an example. 

Or:

https://www.youtube.com/watch?v=3N0klPMNyLY&ab_channel=StatistikinDD

https://r-pkgs.org/vignettes.html

## The **DESCRIPTION** file
This is perhaps the most important document because it contains all of the 
information of the package. For example it includes, the package name (short
and a single word), package title (long and contains more words), version, 
Author information, short description, packages used in the package build,
packages used in the R code, GitHub information and more.

The contents of DESCRIPTION have to be made in a specific way and is expected
to meet certain format requirements. 

For more information see:

https://r-pkgs.org/Metadata.html

But don't worry if something is written incorrectly,
the package checks (discussed later) will tell you. 

## The **NAMESPACE** file
This is automatically generated alongside the .Rd files in **man**. 
You **don't** generally touch the things in NAMESPACE, **However**, you will
need to define how assisting packages used in your functions are imported into
your package. This is used with `import()` and `importFrom()` which also needs
to match how the packages are imported in the .R document in the **R** folder
as well as in the **DESCRIPTION** file.
This can be a bit tricky and look here for a more elaborate explanation:

https://kbroman.org/pkg_primer/pages/depends.html

https://bookdown.dongzhuoer.com/hadley/r-pkgs/namespace.html


## The **README** file
This file gives a brief overview of the package. Typically the description 
section from the DESCRIPTION is included, as well as how to install the 
package. In the end, a section can be included explaining how to cite the
package. 

For a nicer look than a typical .txt document, the README file be made in 
markdown (.md). The README file is automatically displayed on the front page
of the GitHub repository, and therefore making it visually pretty is essential.

## **Optional** files and folders

### The **inst** folder
This folder is necessary if data is included in the package to be used as 
examples for functions. 
Inside the inst folder the **extdata** folder can contain data to be included
in the package. This data can be accessed using

```{r data_access, message=FALSE, results = 'hide', eval = FALSE}
PackageData <- system.file("extdata", "Name_of_file", package = "MyPackage")


```

Then `PackageData` can be used as an object for input into the package functions
This is relevant both in the vignette file, as well as to be used as examples
in the .R document with functions, implemented with `roxygen2`. 

If your package contain **extdata** then you must include a **script** folder in
the **inst** folder. Script includes a markdown (.md) document describing the 
data in the extdata folder. 

Additionally, the **inst** folder can contain a **doc** folder which includes
documentation of the package such as package manual, created with:

```{r manual, message=FALSE, results = 'hide', eval = FALSE}
devtools::build_manual("MyPackage")


```

However, Bioconductor will not allow a **doc** folder in **inst**. 

### The **LICENSE** file
This file describes the user license. Templates exist such as:

https://rdrr.io/github/r-pkgs/usethis/f/inst/templates/license-GPL-3.md

Describing how to generate a LICENSE file. 

# **Creating** the package
Now that you understand what a package contains we will move on to actually 
creating a package. As can be seen above `devtools` is a tool to manipulate
R packages. `devtools` is a very powerful R package made to create and check R 
packages. The following code should be run in your separate 
**creating_MyPackage.R** document.

Installing devtools:

```{r devtools, message=FALSE, results = 'hide', eval = FALSE}

if (!require(devtools)) install.packages('devtools')
library(devtools)

```



Following installing `devtools`, the first you do is run:

```{r creating_folder, message=FALSE, results = 'hide', eval = FALSE}
devtools::create("MyPackage")


```

This will create a folder named **MyPackage** containing, .gitignore, 
.Rbuildignore, DESCRIPTION, MyPackage.Rproj, NAMESPACE files and the R folder. 

The next step is to link this folder to GitHub as explained above. 

Then you create the above mentioned folders and files and then you actually have 
the R package. 

Remember when you have created your MyPackage_functions.R document in **R**, 
then create a **man** folder and run 

```{r creating_documentation, message=FALSE, results = 'hide', eval = FALSE}
devtools::document()


```

To update the **man** folder and the **NAMESPACE** file. 

## **Installing** package from GitHub

At this point you could try and install your package from GitHub using:

```{r intalling_from_GitHub, message=FALSE, results = 'hide', eval = FALSE}
devtools::install_github("usr/MyPackage")


```

Where **usr** is your GitHub username. 

If you have created the necessary files and directories, then your package will 
be installed and you can load the package with:

```{r lib, message=FALSE, results = 'hide', eval = FALSE}
library(MyPackage)


```

Furthermore it is possible to view the `help` pages with:

```{r help, message=FALSE, results = 'hide', eval = FALSE}
?MyPackage::foo()


```

Where `foo()` is an example of a function name contained in the package. 

## **Building** the package

When making R packages we also talk about **builds**. An R package build file is 
a file from which the package can be installed locally. This file is created 
with:

```{r building_the_package, message=FALSE, results = 'hide', eval = FALSE}
devtools::build("MyPackage")


```

## **Checking** the package

A big part of making an R package for publication at bioconductor
is passing 3 `check` functions. These functions will go through all of the files
in your R package folder and check that the files are folders are made 
correctly. In order to acquire these functions run:

```{r access_check1, message=FALSE, results = 'hide', eval = FALSE}
if (!require(devtools)) install.packages('devtools')
library(devtools)

if (!require(BiocManager)) install.packages('BiocManager')
library(BiocManager)


```

`BiocManager` is the Bioconductor manager package, which is used to install 
packages through Bioconductor, including BiocCheck:

```{r access_check2, message=FALSE, results = 'hide', eval = FALSE}

if (!require(BiocCheck)) BiocManager::install("BiocCheck")
library(BiocCheck)

```

`devtools` contain a check function which is always used whereas `BiocCheck`
contain two functions which are Bioconductor specific. 

When something is incorrect in the package the check functions will designate
a **NOTE**, **WARNING**, or an **ERROR** with ascending severity. This means
that `NOTES` primarily are suggestions, however it is recommended to follow the
instructions. `WARNINGS` and `ERRORS` needs to be addressed and fixed. 
`WARNINGS` means that something is not as it is supposed to be, whereas 
`ERRORS` indicate that something is simply not working. 

First run:

```{r devtools_check, message=FALSE, results = 'hide', eval = FALSE}
devtools::check("MyPackage")


```

This needs to be run where the working directory `wd()` is "above" or in the 
root of your package directory. 

Then run:

```{r check_GitClone, message=FALSE, results = 'hide', eval = FALSE}
BiocCheck::BiocCheckGitClone("MyPackage")


```

Which similarly needs to be at the root of your package directory

And finally run

```{r check_BiocCheck, message=FALSE, results = 'hide', eval = FALSE}
BiocCheck::BiocCheck('MyPackage'=TRUE)


```

This needs to be run where the working directory is your package folder

It will take time to address all of the issues with these check functions, 
however it will make your package more streamlined. 

## Package documentation

We have touched upon package documentation throughout this guide. It is also
really important and it ensures reproducibility

When your vignette is done `knit` it and export it as .pdf and .html. 
These files can be kept in your root folder of your package. 

I also recommend using devtools to produce a manual of your functions. 
This manual encompasses all of the help files you have created in the .R folder
Do this by running

```{r build_manual, message=FALSE, results = 'hide', eval = FALSE}
devtools::build_manual("MyPackage")


```

# **Submitting** to Bioconductor
When your package is done and you have passed all 3 check functions the package
is ready to be submitted to Bioconductor. This is done by following this link:

https://github.com/Bioconductor/Contributions/issues

And clicking on "new issue". The only input you have when submitting the package
is your link to the GitHub repository of the package. Therefore, make sure
that you have **updated** the GitHub repository before submitting the package. 

During the review process you'll need to make changes to the package which also
will need to be registered at Bioconductor. Here you'll need to have **git**
installed on your local machine because you'll need to write commands in the 
prompt. How to create this link and to push changes are described in detail
here:

https://contributions.bioconductor.org/git-version-control.html#new-package-workflow


# **Conclusion**

Hopefully this guide has been helpful in your creation of an R package. 
Just a few final notes:

Remember to push to GitHub frequently so your package is updated online

Address all issues returned by the check functions

Document the creation of your package by building the package in a separate .R
file

Frequently uninstall and re-install your package to test whether installation, 
building and activation of the package runs smoothly. 

Create an elaborate vignette. It really helps you understand your code better

Thank you

